ASAH1 rGBM Manuscript ELISA Figure

### 2023-02-06 results/elisa_plasma_proteomics/manuscript_figures_finalized.Rmd
[ADD] finalized ELISA plots, pngs and svgs, and add cox-ph model

<!-- # Helper Functions -->
```{r message=FALSE, warning=FALSE, include=FALSE}
cs <- function(.x) {    # center/scale vector
  print(.x)
  out <- .x - mean(.x)  # center
  out / sd(out)         # scale
}

median_centre_data <- function(train_df){
  values_to = "measure_col"
  measure_col="measure_col"
  train_df %>%
  tibble::rowid_to_column(var = "sample_id_col") %>%
  tidyr::pivot_longer(cols = c(ASAH1, MMP.9, SYNEMIN, GPNMB), names_to = "feature_id_col", values_to = "measure_col") ->
  train_long

@@

@@

#' Calculate the ratio of iGBM to rGBM and rGBM to iGBM for a given data frame
#'
#' @param data_long2 A data frame with columns "gene", "Number.x", "patient_state", "Age at surgery (years)", "Sex", "Tumor localization", "Type of resection", "MGMT promotor methylation", "p53 accumulation", "Ki67-Li", "EGFR vIII", "Treatment", "Time-to-reccurence (days)", "Latency (days)", and "conc".
#' @return A data frame with the calculated ratios.
get_ratio_data <- function(data_long2){
  ## Data wrangling ----
  ### iGBM over rGBM ratio
  data_long2 %>% 
    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        tmp$ratio <- .d_initial$conc / .d_second$conc
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        tmp$ratio <- .d_initial$conc / .d_second$conc
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
    dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
    dplyr::mutate(log2_conc = log2(ratio)) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

  ### rGBM over iGBM
  data_long2 %>% 
    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
    dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        tmp$ratio <- .d_second$conc / .d_initial$conc
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        tmp$ratio <- .d_second$conc / .d_initial$conc
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "post-surgery/pre-surgery"))) %>%
    dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
    dplyr::filter(grepl(";", agg_lab_id)) %>%
    dplyr::mutate(log2_conc = log2(ratio)) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

@@

<!-- # Load Data -->
```{r echo=FALSE, warning=FALSE}
clinical_data <- fread(patient_clinical_data)
patient_longitudinal_mapping_data <- fread(patient_longitudinal_mapping_file)

patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "LAB_ID", all = TRUE)
# patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID <-  patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID.y

elisa_data <- fread(elisa_data_file)

data_long <- merge(elisa_data, patient_meta_data, by = "LAB_ID")
# Remove patient 11, for some reason they are kept
data_long %>%
  dplyr::filter(Number.x!=11) -> data_long

patient_meta_data %>%
  dplyr::filter(!is.na(manifestation_type) & ELISA=="X") %>%
  dplyr::group_by(Number.x) %>%
  dplyr::add_count() %>%
  dplyr::select(`Unique Patient Number`=Number.x, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
  unique() %>% dplyr::ungroup() -> recurrent_patients_meta

# message(sprintf("There are %s patients with intial and recurrent ELISA measurements", nrow(recurrent_patients_meta)))

@@

patient_id_meta <- unique(data_long[, .(LAB_ID, Number.x, collection_type, manifestation_type, patient_state, `Time-to-reccurence (days)`)])
```

@@

comparisons <- matrix(c("Ctrl", "Ctrl", "iGBM", "iGBM", "rGBM", "rGBM"), ncol=2)
methods <- c("perm.test", "perm.test", "paired.perm.test")
# methods <- c("perm.test", "perm.test", "perm.test")

@@

## Build the plot ----
colfunc <- colorRampPalette(c("red","yellow","springgreen","royalblue"))
ttr_categorizer <- function(data) {
  data <- as.numeric(data)
  if (is.na(data)){
    return(NA)
  }
  if (data >= 100 & data < 200){
    return("100-200")
  } else if (data >= 200 & data < 300){
    return("200-300")
  } else if (data >= 300 & data < 400){
    return("300-400")
  } else if (data >= 400 & data < 500){
    return("400-500")
  } else if (data >= 500 & data < 600){
    return("500-600")
  }
}
# ttr_categorizer <- function(data) {
#   data <- as.numeric(data)
#   if (is.na(data)){
#     return(NA)
#   }
#   if (data >= 100 & data < 150){
#     return("100-150")
#   } else if (data >= 150 & data < 200){
#     return("150-200")
#   } else if (data >= 200 & data < 250){
#     return("200-250")
#   } else if (data >= 250 & data < 300){
#     return("250-300")
#   } else if (data >= 300 & data < 350){
#     return("300-350")
#   } else if (data >= 350 & data < 400){
#     return("350-400")
#   } else if (data >= 400 & data < 450){
#     return("400-450")
#   } else if (data >= 450 & data < 500){
#     return("450-500")
#   } else if (data >= 500 & data < 550){
#     return("500-550")
#   } else if (data >= 550 & data < 600){
#     return("550-600")
#   }
# }
# ttr_categorizer <- function(data, split_value) {
#   if (data <= split_value) {
#     return(0)
#   } else {
#     return(1)
#   }
# }
# ttr_categorizer <- function(data) {
#   if (data >= 103 & data < 190){
#     return("early")
#   } else if (data >= 190 & data < 243){
#     return("middle")
#   } else if (data >= 243 & data <= 521){
#     return("late")
#   }
# }
pdata$`ttr_group` <- sapply(as.numeric(pdata$`Time-to-reccurence (days)`), ttr_categorizer)
ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4) +
  # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
  # stat_compare_means(comparisons = my_comparisons, group.by="gene", label.y = c(4, 7, 7), label.x = c(1, 2, 3)) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "iGBM", "rGBM")) +
  # geom_jitter(aes(label=Number.x)) +
  # ggtitle("Initial vs Recurrence") +
  labs(x="", y="log2( Concentration )") +
  facet_wrap(~gene, scales = "free_y") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical",) +
  guides(fill=guide_legend(title="", ncol=3),
         col='none') -> p1_org

p1_org + 
  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(8, 18, 9, 12)),
             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1_org

# Exclude outlier patient number.x 54 
ggpaired(data = pdata[Number.x!=54], id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4) +
  # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
  # stat_compare_means(comparisons = my_comparisons, group.by="gene", label.y = c(4, 7, 7), label.x = c(1, 2, 3)) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "iGBM", "rGBM")) +
  # geom_jitter(aes(label=Number.x)) +
  # ggtitle("Initial vs Recurrence") +
  labs(x="", y="log2( Concentration )") +
  facet_wrap(~gene, scales = "free_y") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical",) +
  guides(fill=guide_legend(title="", ncol=3),
         col='none') -> p1

p1 + 
  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(7, 15, 8, 11)),
             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1
# p1

## Average Conc per ttr
pdata %>%
  dplyr::filter(Manifestation!="Ctrl") %>%
  dplyr::select(gene, Number.x, Manifestation, log2_conc, ttr_group) %>%
  dplyr::mutate(Manifestation = plyr::mapvalues(Manifestation, from=c("iGBM", "rGBM"), to=c("initial\nGBM", "recurrent\nGBM"))) %>%
  dplyr::group_by(gene, Manifestation, ttr_group) %>%
  dplyr::mutate(avg_log2_conc_per_mani_ttr = median(log2_conc, na.rm = T)) -> pdata_ttr

p1_avg_ttr <- ggplot(data = pdata_ttr, aes(x=Manifestation, y=log2_conc)) +
  geom_boxplot() +
  geom_line(aes(x=Manifestation, y=avg_log2_conc_per_mani_ttr, col=ttr_group, group=ttr_group), inherit.aes = FALSE) +
  labs(x="", y="log2( Concentration )") +
  facet_wrap(~gene, scales = "free_y") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical",) +
  guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
         col=guide_legend(title="Avg. log2 ( concentration )\nTime to Recurrence"))

@@

## Time to Recurrence ----
train_df_Sub <- pdata %>%
  dplyr::filter( !(Number.x %in% c("32", "33"))) %>%
  dplyr::filter(Manifestation!="Ctrl") %>%
  dplyr::select(Number.x, Time.to.reccurence..days.=`Time-to-reccurence (days)`, class=Manifestation, gene, log2_conc) %>%
  # dplyr::mutate(Time.to.reccurence..days. = ifelse(class=="iGBM", 0, Time.to.reccurence..days.)) %>%
  # dplyr::mutate(class = ifelse(class=="iGBM", 0, 1) ) %>%
  dplyr::rowwise() %>%
  # dplyr::mutate(ttr_group = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 253.5)) %>%
  # dplyr::select(-Time.to.reccurence..days.) %>%
  dplyr::mutate(gene=make.names(gene)) %>%
  tidyr::pivot_wider(id_cols =c("Number.x", "Time.to.reccurence..days.", "class"), names_from = "gene", values_from = "log2_conc") %>%
  # tidyr::pivot_wider(id_cols = c("Number.x", "Time.to.reccurence..days."), names_from = c("class"), values_from = c("ASAH1", "SYNEMIN", "GPNMB", "MMP.9")) %>%
  # dplyr::mutate(ASAH1 = (ASAH1_rGBM+1) /( ASAH1_iGBM+1),
  #               SYNEMIN = (SYNEMIN_rGBM+1)/(SYNEMIN_iGBM+1),
  #               GPNMB = (GPNMB_rGBM+1)/(GPNMB_iGBM+1),
  #               MMP.9 = (MMP.9_rGBM+1)/(MMP.9_iGBM+1)) %>%
  # dplyr::select(Number.x, Time.to.reccurence..days., ASAH1, SYNEMIN, GPNMB, MMP.9 ) %>% 
  # unique() %>%
  dplyr::arrange(Time.to.reccurence..days.)
train_df_Sub <- na.omit(train_df_Sub)
train_df_Sub$Number.x <- as.factor(train_df_Sub$Number.x)

train_df_Sub %>%
  dplyr::group_by(Number.x, Time.to.reccurence..days.) %>%
  dplyr::mutate(ASAH1_delta = ASAH1[class=="rGBM"] - ASAH1[class=="iGBM"],
                SYNEMIN_delta = SYNEMIN[class=="rGBM"] - SYNEMIN[class=="iGBM"],
                GPNMB_delta = GPNMB[class=="rGBM"] - GPNMB[class=="iGBM"],
                MMP.9_delta = MMP.9[class=="rGBM"] - MMP.9[class=="iGBM"]) %>%
  dplyr::rowwise() %>%
  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 207.5)) %>%
  dplyr::select(Number.x, Time.to.reccurence..days., ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta) %>%
  unique() -> ttr_delta

# ttr_delta$Time.to.reccurence..days. <- as.numeric(as.factor(ttr_delta$Time.to.reccurence..days.))
lm_ttr <- lm(Time.to.reccurence..days. ~ ASAH1_delta + SYNEMIN_delta + GPNMB_delta + MMP.9_delta, data=ttr_delta )
summary(lm_ttr)
anova(lm_ttr)

power_calc <- pwr.f2.test(u = 4, f2 = sqrt(summary(lm_ttr)$adj.r.squared), sig.level = 0.05, power = 0.8)
print(sprintf("The estimated sample size needed for 80%% power is: %s", round(power_calc$v) + 5))

ttr_categorizer <- function(data) {
  if (data >= 103 & data < 190){
    return("early")
  } else if (data >= 190 & data < 243){
    return("middle")
  } else if (data >= 243 & data <= 521){
    return("late")
  }
}

train_df_Sub %>%
  dplyr::group_by(Number.x, Time.to.reccurence..days.) %>%
  dplyr::mutate(ASAH1_delta = ASAH1[class=="rGBM"] - ASAH1[class=="iGBM"],
                SYNEMIN_delta = SYNEMIN[class=="rGBM"] - SYNEMIN[class=="iGBM"],
                GPNMB_delta = GPNMB[class=="rGBM"] - GPNMB[class=="iGBM"],
                MMP.9_delta = MMP.9[class=="rGBM"] - MMP.9[class=="iGBM"]) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 207.5)) %>%
  dplyr::select(Number.x, Time.to.reccurence..days., ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta) %>%
  unique() -> ttr_delta

>>>>

<!-- # Helper Functions -->
```{r message=FALSE, warning=FALSE, include=FALSE}
#' median_centre_data
#'
#' @param train_df  (data.frame) A dataframe that contains labels and features to run through prediction models. Has to have the following structure:
#' \item{class}{(character) sample/observation label. i.e. ctrl, iGBM, or rGBM sample.}
#' \item{feature_1}{(numeric} a specific feature measurement such as protein ASAH1)}
#' \item{feature_n}{(numeric) another specific feature.}
#' @param pred_col_prob (character) A character corresponding to the class of probabilities you want to extract for. i.e. "GBM"
#' @param multiclass (logical) Run predictions using multiclass modelling and prediction if more than 2 classes.
#' 
#' @return a data.frame with columns sample_id_col, class, and the measures normalized by median
#' 
#' This function takes in a train data.frame and returns the data after normalizing the measures 
#' by subtracting the median of each sample from each value and then adding the median of the whole data.
#' The input data.frame is also pivoted for easier analysis.
median_centre_data <- function(train_df){
  values_to = "measure_col"
  measure_col="measure_col"
  train_df %>%
  tibble::rowid_to_column(var = "sample_id_col") %>%
  tidyr::pivot_longer(cols = c(ASAH1, MMP.9, SYNEMIN, GPNMB), names_to = "feature_id_col", values_to = "measure_col") ->
  train_long

@@

median_centre_recurrent_data <- function(data_long){
  data_long[patient_state==1] %>%
    dplyr::group_by(gene, Number.x) %>%
    dplyr::add_count() %>%
    dplyr::filter(n==2) %>% 
    dplyr::ungroup() -> tmp_select

  data_long_tmp <- merge(tmp_select[, c("Number.x", "gene", "patient_state", "LAB_ID", "collection_type", "manifestation_type")], tmp_median_centred, by=c("Number.x", "manifestation_type", "gene"), all.y=T)
  return(data_long_tmp)
}

@@

#' Calculate the ratio of iGBM to rGBM and rGBM to iGBM for a given data frame
#'
#' @param data_long2 A data frame with columns "gene", "Number.x", "patient_state", "Age at surgery (years)", "Sex", "Tumor localization", "Type of resection", "MGMT promotor methylation", "p53 accumulation", "Ki67-Li", "EGFR vIII", "Treatment", "Time-to-reccurence (days)", "Latency (days)", and "conc".
#' @return A data frame with the calculated ratios.
get_ratio_data <- function(data_long2, already_log_transformed=F){
  ## Data wrangling ----
  ### iGBM over rGBM ratio
  data_long2 %>% 
    dplyr::rowwise() %>%
    dplyr::mutate(conc = ifelse(already_log_transformed==T, conc, conc+1)) %>% # Add one to avoid cases of 0 division
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        if (!already_log_transformed){
          tmp$ratio <- .d_initial$conc / .d_second$conc
        } else {
          tmp$ratio <- .d_initial$conc - .d_second$conc
        }
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        if (!already_log_transformed){
          tmp$ratio <- .d_initial$conc / .d_second$conc
        } else {
          tmp$ratio <- .d_initial$conc - .d_second$conc
        }
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
    dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(log2_conc = ifelse(already_log_transformed==T, ratio, log2(ratio))) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

  ### rGBM over iGBM
  data_long2 %>% 
    dplyr::rowwise() %>%
    dplyr::mutate(conc = ifelse(already_log_transformed==T, conc, conc+1)) %>% # Add one to avoid cases of 0 division
    dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        if (!already_log_transformed){
          tmp$ratio <- .d_second$conc / .d_initial$conc
        } else {
          tmp$ratio <- .d_second$conc - .d_initial$conc
        }
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        if (!already_log_transformed){
          tmp$ratio <- .d_second$conc / .d_initial$conc
        } else {
          tmp$ratio <- .d_second$conc - .d_initial$conc
        }
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "post-surgery/pre-surgery"))) %>%
    dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
    dplyr::filter(grepl(";", agg_lab_id)) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(log2_conc = ifelse(already_log_transformed==T, ratio, log2(ratio))) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

@@

<!-- # Load Data -->
```{r echo=FALSE, warning=FALSE}
# Get patient meta data ----
clinical_data <- fread(patient_clinical_data)
patient_longitudinal_mapping_data <- fread(patient_longitudinal_mapping_file)
patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "LAB_ID", all = TRUE)

@@

# Load actual ELISA data ----
elisa_data <- fread(elisa_data_file)

# Merge ELISA data with patient meta data ----
data_long <- merge(elisa_data, patient_meta_data, by = "LAB_ID")
```

@@

comparisons <- matrix(c("Ctrl", "Ctrl", "iGBM", "iGBM", "rGBM", "rGBM"), ncol=2)
methods <- c("perm.test", "perm.test", "paired.perm.test")

@@

## Build the plot ----

ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4, notch = TRUE) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "iGBM", "rGBM")) +
  # geom_jitter(aes(label=Number.x)) +
  # ggtitle("Initial vs Recurrence") +
  labs(x="", y="log2( Concentration )") +
  facet_wrap(~gene, nrow = 2, scales = "free_y") +
  theme_classic() +
  theme(panel.background = element_rect(fill="white"),
        # panel.grid.major.y = element_line(colour = "black"), 
        plot.title = element_text(hjust = 0.5, size=17),
        # axis.line = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical",) +
  guides(fill='none',
         col='none') -> p1_org

dat <- ggplot_build(p1_org)$data[[1]]
dat <- cbind(dat, gene=sort(unique(pdata$gene))[dat$PANEL])

p1_org <- p1_org + geom_segment(data=dat, 
                 aes(x=xmin, xend=xmax, 
                    y=middle, yend=middle), 
                 colour="white", size=1)

p1_org + 
  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(6, 16, 8, 11)),
             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1_org

# Exclude outlier patient number.x 54 
ggpaired(data = pdata[Number.x!=54], id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4) +
  # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
  # stat_compare_means(comparisons = my_comparisons, group.by="gene", label.y = c(4, 7, 7), label.x = c(1, 2, 3)) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "iGBM", "rGBM")) +
  # geom_jitter(aes(label=Number.x)) +
  # ggtitle("Initial vs Recurrence") +
  labs(x="", y="log2( Concentration )") +
  facet_wrap(~gene, scales = "free_y") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical",) +
  guides(fill="none",
         col='none') -> p1

dat <- ggplot_build(p1)$data[[1]]
dat <- cbind(dat, gene=sort(unique(pdata$gene))[dat$PANEL])

p1 <- p1 + geom_segment(data=dat, 
                 aes(x=xmin, xend=xmax, 
                    y=middle, yend=middle), 
                 colour="white", size=1)

p1 + 
  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(6, 15, 8, 11)),
             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1

@@

## Save figure ----
main_fig_a <- p1
# subfigure_6
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.png", plot = main_fig_a, width = 11, height = 9, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.svg", plot = main_fig_a, width = 11, height = 9, dpi = 200)

supp_figure_1_a <- p1_org

ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_1_a_avg.png", plot = supp_figure_1_a, width = 11, height = 9, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_1_a_avg.svg", plot = supp_figure_1_a, width = 11, height = 9, dpi = 200)
```

<!-- # Latency and TTR Data for KM plots-->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
# LAtency data ----
latency_data_file <- "../../data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv"
latency_data <- fread(latency_data_file)
# Clean names
latency_data <- rstatix::make_clean_names(latency_data)
latency_data %>%
  dplyr::mutate(ASAH1.ng.ml = log2(ASAH1.ng.ml+1),
                SYNM.ng.ml = log2(SYNM.ng.ml+1),
                GPNMB.ng.ml = log2(GPNMB.ng.ml+1),
                MMP.9.ng.ml = log2(MMP.9.ng.ml+1)) %>%
  dplyr::mutate(ASAH1 = ifelse(ASAH1.ng.ml>=mean(ASAH1.ng.ml), "high", "low"),
                SYNM = ifelse(SYNM.ng.ml>=mean(SYNM.ng.ml), "high", "low"),
                GPNMB = ifelse(GPNMB.ng.ml>=mean(GPNMB.ng.ml), "high", "low"),
                MMP9 = ifelse(MMP.9.ng.ml>=mean(MMP.9.ng.ml, na.rm=T), "high", "low")) -> latency_data

### ASAH1 ----
fit <- survfit(Surv(Latency..days., status) ~ ASAH1, data = latency_data)
print(fit)
n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$ASAH1)
# Change color, linetype by strata, risk.table color by strata
surv_p1 <- ggsurvplot(fit,
          pval = TRUE, pval.coord = c(1600, 0.85), conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2",
          legend.title = "",
          legend.labs = c(sprintf("ASAH1=high\n(n=%s)", n_data[1]), sprintf("ASAH1=low\n(n=%s)", n_data[2]))
          ) +
  labs(x="Latency Time (Days)")

surv_p1 <- surv_p1$plot +
  theme(axis.title = element_text(size=16),
        axis.text = element_text(size=15),
        legend.text = element_text(size=12),
        legend.spacing.x = unit(0.25, 'cm'),
        legend.position = c(0.8, 0.9))

ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_1.png", plot = surv_p1, width = 11, height = 8.5, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_1.svg", plot = surv_p1, width = 11, height = 8.5, dpi = 200)

### SYNM ----
fit <- survfit(Surv(Latency..days., status) ~ SYNM, data = latency_data)
print(fit)
n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$SYNM)
# Change color, linetype by strata, risk.table color by strata
surv_p2 <- ggsurvplot(fit,
          pval = TRUE, pval.coord = c(1600, 0.85), conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2",
          legend.title = "",
          legend.labs = c(sprintf("SYNM=high\n(n=%s)", n_data[1]), sprintf("SYNM=low\n(n=%s)", n_data[2]))
          ) +
  labs(x="Latency Time (Days)")

surv_p2 <- surv_p2$plot +
  theme(axis.title = element_text(size=16),
        axis.text = element_text(size=15),
        legend.text = element_text(size=12),
        legend.spacing.x = unit(0.25, 'cm'),
        legend.position = c(0.8, 0.9))

ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_2.png", plot = surv_p2, width = 11, height = 8.5, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_2.svg", plot = surv_p2, width = 11, height = 8.5, dpi = 200)
